---
title: "Assignment 1b"
author: "Gian Carlo Diluvi"
date: "November 2023"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I'll code the [LogSumExp trick](https://en.wikipedia.org/wiki/LogSumExp).
This is a useful function to have around&mdash;in my research,
I always have a script with auxiliary functions and this is one of them.
The idea is that $\log\sum_{n=1}^N \exp(x_n)$
can be prone to underflow if any $x_n$ is very negative, 
say in the order of $10^{-10}$.
This seems niche but is actually pretty common when dealing with, e.g.,
mixture models.
In that setting, $x_n$ will correspond to a log pdf,
which can have arbitrarily small values.
The LogSumExp trick refers to the fact that
$$
  \log\sum_{n=1}^N \exp(x_n)
  = x^\star + \log\sum_{n=1}^N \exp(x_n-x^\star)
$$
for any $x^\star$. Taking $x^\star = \max(x_1,\dots,x_N)$
guarantees that $\exp(x_n-x^\star)\leq 1$,
which makes the resulting operation more numerically stable.

```{r logsumexp, include=TRUE}
#' LogSumExp
#' 
#' Compute log(sum(exp(x))) in a numerically-stable fashion
#'
#' @param x numeric, ideally a vector 
#' (name rationale: usual name for variable in my research area is x)
#' Ellipses added for max and exp functions, e.g. to remove NAs from x
#' 
#' @return numeric, single number
logsumexp <- function(x,...){
  x_max <- max(x,...)
  result <- x_max + log(sum(exp(x-x_max,...)))
  return(result)
}

```


## Some examples

```{r}
my_x = c(-1e5,-1e12)
logsumexp(my_x)
log(sum(exp(my_x)))
```

